Modeling and analyzing the Corona-virus warning app with the Isabelle Infrastructure framework

### 2020-07-08 paper.tex
added the conclusion subjection 7.2

\subsubsection{Refinement}
An additional feature that has been integrated into the Isabelle Infrastructure
framework motivated by security engineering formal specifications for IoT healthcare
system is an extension of the formal specification process introducing
refinement of Kripke structures \cite{kam:19a,kam:20a}. It refines a system model based on a 
formal definition of a combination of trace refinement and structural 
refinement (or datatype refinement). The definition allows to prove property preservation results 
crucial for an iterative development process.
The refinements of the system specification  can be interleaved with attack 
analysis while security properties can be proved in Isabelle. In each iteration
security qualities are accumulated while continuously attack trees scrutinize
the design.
One of the contributions of this paper is to explore different concepts of refinement:
the formal expression of refinemnet, enables to pin down (i.e. exemplify) different concepts
of refinement (data refinement, action refinement, trace refinement (aka spec refinement) and
combinations thereof with concrete attack scenarios.

@@

\subsection{Relevance of the approach}

>>>>

\subsubsection{Refinement}
An additional feature that has been integrated into the Isabelle Infrastructure
framework motivated by security engineering formal specifications for IoT healthcare
system is an extension of the formal specification process introducing
refinement of Kripke structures \cite{kam:19a,kam:20a}. It refines a system model based on a 
formal definition of a combination of trace refinement and structural 
refinement (or datatype refinement). The definition allows to prove property preservation results 
crucial for an iterative development process.
The refinements of the system specification  can be interleaved with attack 
analysis while security properties can be proved in Isabelle. In each iteration
security qualities are accumulated while continuously attack trees scrutinize
the design.
One of the contributions of this paper is to explore different concepts of refinement:
the formal expression of refinement, enables to pin down (i.e. exemplify) different concepts
of refinement (data refinement, action refinement, trace refinement (aka spec refinement) and
combinations thereof with concrete attack scenarios.

@@

\subsection{Summary and discussion of relevance of the approach}
We can establish in our formal framework an attack that even a system using changing
Ephemeral ids can be broken if the attacker physically follows a victim. This is a basic attack on
anonymity: a user's connection between his Ephemeral Ids and personal details (Iphone MAC or name)
is revealed to the attacker. The protection goal of privacy is thereby destroyed.

When establishing the attack we start from a simplistic scenario that does not use Ephemeral Ids but fixed ids.
In this (over)simplified model the attack is established. We then define a formal Refinement calculus for
the Isabelle Infrastructure framework to refine the attack to a system with Ephemeral Ids that change in fixed
time intervals obfuscating the relationship between user and his pseudonym\footnote{We identify the smartphone
  and the user which might be also recognized by his appearance (face)}.

Now, the refinement shows that although the Ephemeral Ids change regularly the same attack that has been
identified on the very abstract level (fixed ids) persists.
The refinement allows refining the datatype (Id $\mapsto$ EphId) but also delivers the usual trace
refinement (behaviours of the refined systems are a subset of the traces of the abstract system).
This persistence of the attack precisely shows which part of the system behaviour is responsible for the
attack. In a second refinement step using action refinement based on this insight from the (repeated)
attack, we can exclude the dangerous attack trace\footnote{Implication on the actual protection goals
  that are endangered by the attack on privacy are discussed in the larger context of the Corona-warning
  app (see previous Section)}.

Clearly, the abstract attack we establish is obvious on an informal level but the persistence of the attack
on the refined system is less obvious. The remedy by a second refinement step is an evident restriction of
the system behaviour which gives a clear specification of a system secured against this attack. The use of
formal methods therefore lies not in the discovery of an obvious attack on a simplified system but in showing
how a formal specification including security refinement can lead to a stepwise improvement that is accompanied
by formal proof in the Isabelle Infrastructure framework. The solution to exclude the attack in the second refinement
step binds the action move together with a put action. %which effectively represents an action refinement.
This shows that besides datatype refinement and trace refinement our Refinement calculus also entails action
refinement. This action refinement is implemented implicitly by changing the effects of the actions
in the semantic state transition relation. In future work, we could think about making the action refinement
more explicit by considering a relationship between semantic rules or by refining the refinement notion to
a more explicit layer of protocol steps -- similar to what has been done in previous applications of the
Isabelle Infrastructure framework for example to Auction protocols \cite{kkp:17} or the Quantum Key Distribution \cite{kam:19c}.

The security refinement in general might seem pointless, as in the first step of refinement the attack
persists and even though the second refinement gets rid of this specific attack, it doesn't exclude the
reachability of the attack goal altogether (if the attacker Eve gets Bob on his own in a location she can
map all used EphIDs to him).
However, the refinement makes the system relatively more secure in that for a larger number of traces the abstract
attack does not work anymore\footnote{Adding probabilities as in \cite{kam:19c} enables quantifying this.}. 
It is important to emphasize that security refinement is a cyclic process that
improves security but does not usually terminate (like a loop) with a fixed point of 100\% secure system.
In a refined model new detail may give rise to new attack possibilities. These additional attacks can be
identified using the Attack Tree calculus and trigger further refinement steps.
